### 2017-07-07 motivation.tex

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as follows:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariants, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The following two sections,
however, first focus on laying the theoretical foundations for
reasoning about weak isolation.

>>>>

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$\T_U$) is the following function on sets\footnote{$\bind$ has higher
precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
           {\emptyset}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$. 
The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\Delta,\delta) =
\stl \cup F(\stg)$, where $F$ is a function that returns the set of
records added by the transformer to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for $\T_U$
and $\T_I$ shown above. The state transformed of the loop body in
Fig.~\ref{fig:new} can be expressed as the following composition of
$\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
Thus, state transformer inference is completely mechanical. The
advantage of inferring such transformers is that we can now make use
of a semantics-preserving translation from the domain of sets equipped
with $\bind$ to first-order logic, allowing us to leverage SMT solvers
for automatic proofs without having to infer loop invariants, or
provide intermediate assertions.  Sec.~\ref{sec:inference} describes
this translation. In the exposition thus far, we assumed $\Delta$
remains invaraint, which is clearly not the case when we admit
concurrency.  Necessary concurrency extensions of the state
transformer semantics to deal with interference is also covered in
Sec.~\ref{sec:inference}.  The following two sections, however, first
focus on laying the theoretical foundations for reasoning about weak
isolation.
